```r
---
title: "Churn Analysis - EDA"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

This report presents exploratory data analysis for customer churn. The goal is to identify patterns and insights related to customer churn that can inform predictive modeling.

## 2. Data Loading

Load the cleaned customer data and transaction data for analysis.

```{r load-data}
library(tidyverse)
library(ggplot2)
library(readr)
library(dplyr)
library(gridExtra)
library(corrplot)
library(scales)

# Load the data
customer_data <- read_csv("data/processed/cleaned_customer_data.csv")
transaction_data <- read_csv("data/raw/transaction_data.csv")

# Preview data
head(customer_data)
head(transaction_data)
```

## 3. Data Overview

Check the structure and summary of the data to understand its variables and their distributions.

```{r data-overview}
# Structure of the data
str(customer_data)

# Summary of the data
summary(customer_data)

# Checking for missing values
sapply(customer_data, function(x) sum(is.na(x)))

# Checking for duplicated records
sum(duplicated(customer_data))
```

## 4. Distribution of Key Variables

Visualize the distribution of key variables such as customer tenure, monthly charges, and total charges.

```{r distribution-plots}
# Tenure Distribution
p1 <- ggplot(customer_data, aes(x = tenure)) + 
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  labs(title = "Distribution of Tenure", x = "Tenure (Months)", y = "Frequency")

# Monthly Charges Distribution
p2 <- ggplot(customer_data, aes(x = MonthlyCharges)) + 
  geom_histogram(bins = 30, fill = "green", alpha = 0.7) +
  labs(title = "Distribution of Monthly Charges", x = "Monthly Charges", y = "Frequency")

# Total Charges Distribution
p3 <- ggplot(customer_data, aes(x = TotalCharges)) + 
  geom_histogram(bins = 30, fill = "purple", alpha = 0.7) +
  labs(title = "Distribution of Total Charges", x = "Total Charges", y = "Frequency")

# Display the plots together
grid.arrange(p1, p2, p3, ncol = 3)
```

## 5. Churn Rate Analysis

Analyze the churn rate and compare churned vs non-churned customers based on features.

```{r churn-rate}
# Proportion of churn
churn_rate <- customer_data %>% 
  group_by(Churn) %>% 
  summarise(count = n()) %>% 
  mutate(percentage = count / sum(count) * 100)

# Visualizing churn rate
ggplot(churn_rate, aes(x = Churn, y = percentage, fill = Churn)) + 
  geom_bar(stat = "identity", alpha = 0.7) + 
  labs(title = "Churn Rate", x = "Churn", y = "Percentage") +
  scale_y_continuous(labels = percent_format())
```

## 6. Churn by Customer Segments

Analyze churn based on demographic and customer-related factors like gender, internet service type, and contract type.

```{r churn-segments}
# Churn by Gender
p1 <- ggplot(customer_data, aes(x = gender, fill = Churn)) +
  geom_bar(position = "fill", alpha = 0.7) + 
  scale_y_continuous(labels = percent_format()) + 
  labs(title = "Churn by Gender", y = "Percentage")

# Churn by Internet Service
p2 <- ggplot(customer_data, aes(x = InternetService, fill = Churn)) + 
  geom_bar(position = "fill", alpha = 0.7) +
  scale_y_continuous(labels = percent_format()) + 
  labs(title = "Churn by Internet Service", y = "Percentage")

# Churn by Contract Type
p3 <- ggplot(customer_data, aes(x = Contract, fill = Churn)) + 
  geom_bar(position = "fill", alpha = 0.7) + 
  scale_y_continuous(labels = percent_format()) + 
  labs(title = "Churn by Contract Type", y = "Percentage")

# Display the plots together
grid.arrange(p1, p2, p3, ncol = 3)
```

## 7. Correlation Analysis

Perform correlation analysis to explore relationships between numeric variables and churn.

```{r correlation-analysis}
# Convert Churn to numeric (1 for Yes, 0 for No)
customer_data$Churn <- ifelse(customer_data$Churn == "Yes", 1, 0)

# Select numeric variables
numeric_vars <- customer_data %>%
  select(tenure, MonthlyCharges, TotalCharges, Churn)

# Compute correlation matrix
correlation_matrix <- cor(numeric_vars, use = "complete.obs")

# Visualize correlation matrix
corrplot(correlation_matrix, method = "color", type = "lower", 
         tl.col = "black", tl.srt = 45, 
         addCoef.col = "black", number.cex = 0.7, 
         col = colorRampPalette(c("red", "white", "blue"))(200))
```

## 8. Churn by Payment and Contract Details

Analyze how payment methods and contract lengths impact churn rates.

```{r payment-contract-churn}
# Churn by Payment Method
p1 <- ggplot(customer_data, aes(x = PaymentMethod, fill = Churn)) + 
  geom_bar(position = "fill", alpha = 0.7) + 
  scale_y_continuous(labels = percent_format()) + 
  labs(title = "Churn by Payment Method", y = "Percentage")

# Churn by Contract Length
p2 <- ggplot(customer_data, aes(x = Contract, fill = Churn)) + 
  geom_bar(position = "fill", alpha = 0.7) + 
  scale_y_continuous(labels = percent_format()) + 
  labs(title = "Churn by Contract Length", y = "Percentage")

# Display the plots together
grid.arrange(p1, p2, ncol = 2)
```

## 9. Customer Lifetime Value (CLV) and Churn

Explore the relationship between CLV (calculated from transaction data) and churn.

```{r clv-churn}
# Compute Customer Lifetime Value (CLV)
transaction_summary <- transaction_data %>%
  group_by(CustomerID) %>%
  summarise(TotalSpend = sum(SpendAmount))

# Merge CLV with customer data
customer_clv <- customer_data %>%
  left_join(transaction_summary, by = "CustomerID")

# Visualizing CLV vs. Churn
ggplot(customer_clv, aes(x = TotalSpend, fill = Churn)) + 
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  labs(title = "Customer Lifetime Value (CLV) vs. Churn", x = "Total Spend (CLV)", y = "Frequency")
```